\mutagen: Coverage-Guided, Property-Based Testing using Exhaustive Structure-Preserving Mutations

### 2022-03-02 papers/fse22/main.tex
Compressing Heuristics section

This heuristic first tackles the issue of enqueuing new interesting mutation
candidates at the end of (possibly) long queues of not-so-interesting previously
executed ones.

%% \fuzzchick uses AFL instrumentation under the hood, which in turn uses an
%% \emph{edge coverage} criterion to distinguish novel executions and to assign
%% each mutation candidate a given energy.
%% %
Execution traces in \mutagen represent the specific \emph{path} in the code
taken by the program, as opposed to just the (unordered) set of edges traversed
in the control-flow graph (CFG) used by default by most coverage-guided fuzzers.
%
Using this criterion lets us gather precise information from each new execution.
%
%% To keep track of the execution path of each test, our tool keeps an internal
%% prefix tree that records every new entry by traversing this data structure and
%% inserting new nodes after a new path prefix gets discovered.
%% %
In particular, we are interested in the \emph{depth} where each new execution
branches from already seen ones.
%
Our assumption is that test case executions that branch at shallower depths from
the ones already executed are more likely to discover new portions of the code
under test, and hence we want to prioritize them.

@@

In this light, every time we insert a new execution path into the internal trace
log (TLog), we calculate the number of \emph{new nodes} that were executed, as
well as the \emph{branching depth} where they got inserted.
%
The former is used to distinguish interesting test cases (whether or not new
nodes were inserted), whereas the latter is used by this heuristic to schedule
mutation candidates.
%
Fig. \ref{fig:mutagen:tracelog} illustrates this idea, inserting two execution
traces (one after another) into a trace log that initially contains a single
execution path.
%
The second insertion (with trace $1 \rightarrow 2 \rightarrow 6 \rightarrow 7$)
branches at a shallower depth than the first one (2 vs. 3), and so its
corresponding test case is given a higher priority.

@@

To pick the next test case, we retrieve the one with the highest priority (the
smallest branching depth).
%
Then, when we find a new interesting test case, it gets enqueued at the
\emph{beginning} of the queue at its corresponding priority.
%
This allows the testing loop to jump immediately onto mutating new interesting
test cases as soon as they are found (even at the same priority), and to jump
back to previous test cases as soon as their mutants become less novel.

@@

As introduced in \S \ref{sec:mutagen}, our tool is parameterized by the
number of random mutations to be generated over each mutable subexpression
defined using a random mutant, e.g., numeric values.
%
But, how many random mutations should we use? A single one? A few tenths? A few
hundred?
%% %
%% Using too little can put finding bugs at risk.
%% %
%% For instance, when the system under test branches based on numeric values, we
%% should make sure that we set enough random mutations to test each branch on a
%% reasonable basis.
%% %
%% Using too many, on the other hand, can degrade the performance of the testing
%% loop, as it will spend too much time producing uninteresting mutations.
%% %
%% This can happen for instance if the subexpressions defined using random mutants
%% are only used as payloads, and their value does not affect the execution in any
%% way.
%
Answering this question precisely is not an easy task, and this second heuristic
aims to tackle this issue.

We found that, the smaller the number of random mutations we set, the easier it
is for the trace log that records executions to start getting saturated, i.e.,
when interesting test cases stop getting discovered or are discovered very
sporadically.
%
We realized that we can use this information to automatically optimize the
number of random mutations used by our tool.
%
This idea is described in Algorithm \ref{algo:mutagen:reset}.
%
The process is simple:
%
\begin{inparaenum}
  \item we start the testing loop with the number of random mutations set to
    one, and
  %
  \item each time we find that a test is not interesting (i.e. boring), we
    increment a counter.
  %
  Then,
  \item if we have not produced any interesting test case after a certain
    threshold (1000 tests seems to be a reasonable value in practice), we
    increment the number of random mutations and the threshold by twice the
    current amount.
    %
    Additionally, each time this happens we also reset the trace log, so
    interesting test cases found on a previous iteration can be found and
    enqueued for mutation again --- this time with a higher effort dedicated to
    producing random mutations.
\end{inparaenum}

Notably, this heuristic can be particularly useful when the execution of the
system under test depends heavily on some invariant encoded by a subexpression
of its input that get transformed using a random mutant (e.g, the number of
pixels declared by the header of an image matching the size of its actual
payload).
%
There, starting with a single random mutation will quickly saturate the trace
space with discarded (invalid) tests, and this heuristic will continuously
increase the random mutations parameter until the random mutant is given enough
chances to randomly generate the value that satisfies the required invariant and
makes the overall test case valid.

>>>>

This first heuristic tackles the issue of enqueuing new interesting test cases
at the end of (possibly) long queues of not-so-interesting previously executed
ones.

%% \fuzzchick uses AFL instrumentation under the hood, which in turn uses an
%% \emph{edge coverage} criterion to distinguish novel executions and to assign
%% each mutation candidate a given energy.
%% %
Execution traces in \mutagen represent the specific \emph{path} in the code
taken by the system under test, as opposed to just the unordered set of edges
traversed in the control-flow graph used by default by most coverage-guided
fuzzers.
%
Using this criterion we can gather precise information from each new execution.
%
%% To keep track of the execution path of each test, our tool keeps an internal
%% prefix tree that records every new entry by traversing this data structure and
%% inserting new nodes after a new path prefix gets discovered.
%% %
In particular, we are interested in the \emph{depth} where each new execution
branches from already seen ones.
%
Our assumption is that test case executions branching at shallower depths from
already executed ones are more likely to discover new portions of the system
under test, and hence we want to prioritize them.

@@

Every time we insert a new execution path into the internal trace log (TLog), we
calculate the number of \emph{new nodes} that were executed, as well as the
\emph{branching depth} where they got inserted.
%
The former is used to distinguish interesting test cases (whether or not new
nodes were inserted), whereas the latter is used by this heuristic to prioritize
some test cases over others.
%
Fig. \ref{fig:mutagen:tracelog} illustrates this idea, inserting two execution
traces (one after another) into a trace log that initially contains a single
execution path.
%
The second insertion (with trace $1 \rightarrow 2 \rightarrow 6 \rightarrow 7$)
branches at a shallower depth than the first one (2 vs. 3), and so its
corresponding test case is given a higher priority.

@@

To pick the next test case, we retrieve the first one with the highest priority
(the smallest branching depth).
%
Then, when we find a new interesting test case, it gets enqueued at the
\emph{beginning} of the queue of its corresponding priority.
%
This allows the testing loop to jump immediately onto mutating new interesting
test cases as soon as they are found (even at the same priority), and to jump
back to previous test cases as soon as mutants become less novel.

@@

As introduced in \S \ref{sec:mutagen}, our tool is parameterized by the number
of mutated test cases to be generated for each subexpression of the input test
case associated to random mutant.
%
But, how many random test cases should we sample? A single one? A few tenths? A
few hundred?
%% %
%% Using too little can put finding bugs at risk.
%% %
%% For instance, when the system under test branches based on numeric values, we
%% should make sure that we set enough random mutations to test each branch on a
%% reasonable basis.
%% %
%% Using too many, on the other hand, can degrade the performance of the testing
%% loop, as it will spend too much time producing uninteresting mutations.
%% %
%% This can happen for instance if the subexpressions defined using random mutants
%% are only used as payloads, and their value does not affect the execution in any
%% way.
%
Answering this question precisely can be challenging, so this second heuristic
aims to alleviate the problem.

We found that, the smaller the number of random mutations we initially set, the
easier it is for the trace log that records executions to get saturated, i.e.,
when interesting test cases stop getting discovered or are discovered very
sporadically.
%
We realized that we can use this information to automatically optimize the
number of random mutations used by our tool.
%
This idea is described in Algorithm \ref{algo:mutagen:reset}.
%
The process is as follows:
%
\begin{inparaenum}
  \item we start the testing loop with the number of random mutations set to
    one, and
  %
  \item each time we find that a test is not interesting (i.e. boring), we
    increment a counter.
  %
  Then,
  \item if we have not produced any interesting test case after a certain number
    of tests (1000 tests seems to be a reasonable threshold in practice), we
    increment the number of random mutations and the threshold by twice the
    current amount.
    %
    Additionally, each time this happens we reset the trace log, so interesting
    test cases found on a previous iteration can be found and enqueued for
    mutation again with a higher effort dedicated to sampling mutants.
\end{inparaenum}

Notably, this heuristic can be useful when the execution of the system under
test depends heavily on some invariant encoded by a subexpression of its input
that get transformed using a random mutant (e.g, the number of pixels declared
by the header of an image matching the size of its actual payload).
%
There, starting with a single random mutation will quickly saturate the trace
space with discarded (invalid) tests, and this heuristic will continuously
increase the random mutations parameter until the random mutant is given enough
chances to randomly generate a value that satisfies the required invariant and
makes the overall test case valid.
